The imprint of global magma oceans on exoplanet demographics

### 2023-03-14 src/tex/ms.tex
discuss Kepler, K2, TESS

% Main body
\section{Introduction}
...just dropping material and references for now...
\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
\todo[inline]{``Points for the intro:
\bioverse\\
Transit survey to measure mass/radius of planets\\
Does the HZ have an inner edge?\\
Run-away GH\\
Magma oceans\\
Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman. Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
~\\
In contrast, PLATO and Nautilus are \\
Possibility of follow-up to measureme masses
''}

@@

\subsection{Synthetic planet sample}\label{sec:syn_planets}

Next, we assigned to the stellar sample planetary systems with frequencies, orbital parameters, and bulk properties derived from the \kepler\ mission.
We adopted the planetary occurrence rate model from \citet{Bergsten2022}, which defines the occurrence of small planets in radius and orbital period. Following \citet{Youdin2011a}, their inferred occurrence rate density can be expressed in the form:
\begin{equation}
    \frac{\mathrm{d}^2n}{\mathrm{d}R \, \partial P} = F_0 C_n g(R, P),
\end{equation}
where $F_0$ represents the average number of planets per star and $C_n$ is a normalization constant. The shape function $g(R,P)$ describing the occurrence distribution is the same as in \citet{Bergsten2022}. The functional form and stellar-mass dependent parameters (including a new fit to the sample of \kepler\ planets with hosts below $0.56\,M_{\odot}$) are included in \bioverse\ as the prominent planet-generation model; the original model based on SAG13 occurrence rates is preserved as a legacy feature.

@@

\subsubsection{False positive scenarios}
\todo{what other processes could be confused with a magma ocean signal?\\
If there's no other subsubsection here, make this a (the final?) subsection in ``Discussion''.}
\begin{note}
    Runaway greenhouse climates are not the only physical mechanism that may cause a change in transit radius for a subset of planets.
    Some of the most prominent alternative scenarios involve atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
    Both processes are being traded as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}.
%    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
    Either of them leads to a decrease of planet radius for planets close to their host star, which is distinct from the radius inflation introduced by runaway greenhouse climates.

@@

\subsection{Diagnostic power of near-future exoplanet missions}\label{sec:dis_samplesize}
\todo[inline]{discuss Nautilus}
The fidelity of a future detection or falsification of a runaway greenhouse signature will depend on the significance with which the null hypothesis can be excluded.
As a consequence, instrumentation and survey strategy aiming at probing the runaway greenhouse threshold should aim at maximizing the probability of a true positive detection.

@@

What other planned missions are suited to test the runaway greenhouse hypothesis?
The ongoing \cheops\ mission was designed as a follow-up mission to search for transits of planets discovered with other techniques, in particular with radial velocity measurements~\citep{Benz2021}.
As such, it will provide precise radius constraints on a sample of small planets; however, only a small number of planets with periods \SI{> 50}{\day} are being observed.
This largely limits \cheops' coverage to planets within the runaway greenhouse regime, preventing a detection of the transition.

@@

While not primarily designed to detect transiting planets, the \rst~\citep{Spergel2015} is expected to yield $\sim 10^5$ transiting planets on short orbits and constrain their radii in the course of its mission~\citep{Montet2017}.
In a fraction of the cases, planets as small as \SI{2}{\Rearth} will be detected.
About 200 planets smaller than Neptune are expected to be found around M~dwarfs and thus may reach into the habitable zone.
The \rst\ could thus provide a useful sample to explore the runaway greenhouse transition, albeit with a predominant focus on water-rich (sub-)Neptunes~\citep[e.g.,][]{Pierrehumbert2022}.
\todo[inline]{what about TESS?}

Looking further ahead, the Nautilus Space Observatory concept~\citep{Apai2019} represents a statistical mission able to provide precise radius measurements of a large sample of small exoplanets.
It employs a swarm of large telescopes using ultralight diffractive-refractive optical elements~\citep{Milster2020} with the primary goal to study the atmospheres of transiting exoplanets.
Operating in an array mode, Nautilus would achieve the equivalent light-collecting area of a \SI{50}{\meter} telescope.
Its expected \SI{1}{\ppm} photometric precision~\citep{Apai2022} would enable precise radius measurements of a large sample, also through a low number of required visits per object.
If realized, Nautilus will be a valuable instrument for characterizing the runaway greenhouse transition.

As for exoplanets missions in their implementation phase, \plato\ overall appears to be the most promising mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.

>>>>

% Main body
\section{Introduction}
...just dropping material and references for now...
\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
\todo[inline]{``Points for the intro:
\bioverse\\
Transit survey to measure mass/radius of planets\\
Does the HZ have an inner edge?\\
Run-away GH\\
Magma oceans\\
Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman. Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
~\\
In contrast, PLATO and \nautilus\ are \\
Possibility of follow-up to measureme masses
''}

@@

\subsection{Synthetic planet sample}\label{sec:syn_planets}
Next, we assigned to the stellar sample planetary systems with frequencies, orbital parameters, and bulk properties derived from the \kepler\ mission.
We adopted the planetary occurrence rate model from~\citet{Bergsten2022}, which defines the occurrence of small planets in radius and orbital period. Following \citet{Youdin2011a}, their inferred occurrence rate density can be expressed in the form:
\begin{equation}
    \frac{\mathrm{d}^2n}{\mathrm{d}R \, \partial P} = F_0 C_n g(R, P),
\end{equation}
where $F_0$ represents the average number of planets per star and $C_n$ is a normalization constant. The shape function $g(R,P)$ describing the occurrence distribution is the same as in \citet{Bergsten2022}. The functional form and stellar-mass dependent parameters (including a new fit to the sample of \kepler\ planets with hosts below $0.56\,M_{\odot}$) are included in \bioverse\ as the prominent planet-generation model; the original model based on SAG13 occurrence rates is preserved as a legacy feature.

@@

\subsubsection{False positive scenarios}\label{sec:dis_falsepositive}
\todo{what other processes could be confused with a magma ocean signal?\\
If there's no other subsubsection here, make this a (the final?) subsection in ``Discussion''.}
\begin{note}
    Runaway greenhouse climates are not the only physical mechanism that may cause a change in transit radius for a subset of planets.
    Some of the most prominent alternative scenarios involve atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
    Both processes are being traded as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}.
%    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
    Either of them leads to a decrease of planet radius for planets close to their host star, which is distinct from the radius inflation introduced by runaway greenhouse climates.

@@

\subsection{Diagnostic power of near-future exoplanet missions}\label{sec:dis_samplesize}
The fidelity of a future detection or falsification of a runaway greenhouse signature will depend on the significance with which the null hypothesis can be excluded.
As a consequence, instrumentation and survey strategy aiming at probing the runaway greenhouse threshold should aim at maximizing the probability of a true positive detection.

@@

What other planned missions are suited to test the runaway greenhouse hypothesis?
\kepler\ and \ktwo\ have contributed a large number of discovered terrestrial-sized planets, but few of them are in the habitable zone and their host stars are typically too faint for RV follow-up~\citep{Dressing2015}.

Similarly, the \tess\ planet sample lacks temperate, small planets around bright host stars~\citep{Ment2023}, as was expected from planet yield calculations~\citep{Barclay2018}.

The ongoing \cheops\ mission was designed as a follow-up mission to search for transits of planets discovered with other techniques, in particular with radial velocity measurements~\citep{Benz2021}.
As such, it will provide precise radius constraints on a sample of small planets; however, only a small number of planets with periods \SI{> 50}{\day} are being observed.
This largely limits \cheops' coverage to planets within the runaway greenhouse regime, preventing a detection of the transition.

@@

While not primarily designed to detect transiting planets, the \rst~\citep{Spergel2015} is expected to yield $\sim 10^5$ transiting planets on short orbits and constrain their radii in the course of its mission~\citep{Montet2017}.
In a fraction of the cases, planets as small as \SI{2}{\Rearth} will be detected.
About 200 planets smaller than Neptune are expected to be found around M~dwarfs and thus may reach into the habitable zone.
The \rst\ could thus provide a useful sample to explore the runaway greenhouse transition, albeit with a predominant focus on water-rich (sub-)Neptunes~\citep[e.g.,][]{Pierrehumbert2022}.

Looking further ahead, the \nautilus\ Space Observatory concept~\citep{Apai2019} represents a statistical mission able to provide precise radius measurements of a large sample of small exoplanets.
It employs a swarm of large telescopes using ultralight diffractive-refractive optical elements~\citep{Milster2020} with the primary goal to study the atmospheres of transiting exoplanets.
Operating in an array mode, \nautilus\ would achieve the equivalent light-collecting area of a \SI{50}{\meter} telescope.
Its expected \SI{1}{\ppm} photometric precision~\citep{Apai2022} would enable precise radius measurements of a large sample, also through a low number of required visits per object.
If realized, \nautilus\ will be a valuable instrument for characterizing the runaway greenhouse transition.

As for exoplanets missions in their implementation phase, \plato\ overall remains to be the most promising mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.
